---
published: false
title: Quick'n'dirty string categorization using the stringdist library
layout: post
---

When filling out forms on the web or into some other database, often an 'other' category allows for the user to enter some sort of free-response text. So the database will end up with two columns, one for the categories the database administrator expected to be most popular and one that stores text entered by the user if they specified 'other' in the first column.

During a recent project it was important for me to identify occupation types stored in a very large database generated from a web enrolment form. For whatever reason, a very large proportion of users selected `other` then entered their occupation as text rather than choosing it from an adaptive form. Many of these `other` responses appeared to match existing categories very closely, so rather than discarding the `other` responses I thought it was worth my time to hack a quick algorithm to sort these responses into existing categories where possible.

The algorithm takes a list of strings `a` and attempts to match each item from `a` against each item in a list of strings `b`. The best match is stored, with the intention that those with a high enough similarity will be classified according to their best match in `b`. In this example, the list `a` is all of the strings from the `other` text responses and the list `b` is all of the strings from the database-specified list of occupations. 

I recommend converting each list to lower-case first.

```{r, echo = T, message = F, warning = F}
require(stringdist)

fuzzy_match <- function(a, b, method = c("jw","dl","lcs","soundex")){  
  k <- data.frame(NA,NA,NA)
  names(k) <- c("score", "a", "b")
  
  for(i in a){
    x0 <- 0
    for(j in b){
      x <- stringsim(i, j, method = method)
      if(x0 < x){
        x0 <- round(x, 3)
        aa <- i
        bb <- j
      }
    }
    k <- rbind(k, c(x0, aa, bb))
  }
  k <- k[-1,]  # get rid of the first [empty] row
}
```

```{r, message = F, warning = F}
k1 <- fuzzy_match(c("apple", "bears", "peasr", "bees", "raeb", "frappe"), c("apple", "bear", "pear", "donkey"), method = "jw")
k2 <- fuzzy_match(c("apple", "bears", "peasr", "bees", "raeb", "frappe"), c("apple", "bear", "pear", "donkey"), method = "dl")
require(knitr)
kable(k1)
kable(k2)
```

Where the `score` exceeds some sensible threshold, rows with an `other` response can then be updated with the new category.